---
title: "R Notebook"
output: html_notebook
---

## Sour Cream

### a)

Analyze the data in a three-way experiment without interactions with acidity as response and starter, batch and position as factors. By using summary command, can you tell whether there is a significant difference between the effects of starter 1 and starter 2 on acidity? Motivate your answer.

```{r}
data <- read.csv("data/cream.txt", sep="", header=TRUE)
summary(data)
```

### b)

Recall that the main interest is in the effect of starter on acidity; factors position and batch represent the block variables. Remove insignificant block variable(s) if there are such, and perform an ANOVA for the resulting “fixed effects” model. Which starter(s) lead to significantly different acidity? Motivate your answer.

```{r}

```

### c)

For the resulting model from c), can we also apply the Friedman test to test whether there is an effect of starter on acidity?

```{r}
friedman.test(data) #?
```

### d)

Repeat c) by performing a mixed effects analysis, modeling the block variable(s) (if there are any) as a random effect by using the function lmer. (You will need to install the R-package lme4, which is not included in the standard distribution of R.) Compare your results to the results found by using the fixed effects model in c). Comment.

```{r}
lmer(data) #?
```


